Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-04 ind_files/03_BD_processes.tex
Aesthetic changes to .cls file - make title bold in certificate

b^{(K)}_i(\mathbf{x}) &= \mu Q_i(\mathbf{x}) + b^{(\textrm{int})}_{i}(\mathbf{x})x_i\\
d^{(K)}_i(\mathbf{x}) &= d^{(\textrm{int})}_i(\mathbf{x})x_i

@@

where $\mu \geq 0$ is a constant describing the mutation rate, $Q_i(\mathbf{x})$ is a real-valued function that describes the birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$, and $b^{(\textrm{int})}_{i}(\mathbf{x})$ and $d^{(\textrm{int})}_{i}(\mathbf{x})$ are real-valued functions that respectively describe the per-capita birth and death rate of type $i$ individuals due to ecological interactions. Our assumptions of the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates. Plugging these definitions into \eqref{nD_det_limit} and writing it down component-wise, we obtain the equation

@@

Given a state $\mathbf{x}(t)$, we can now compute the total scaled population size, and the frequency of each type in the population as:

@@

We can also define the mean of any type level quantity $f$ in the population as

@@

Where we have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. The first term of \eqref{nD_replicator_mutator} describes changes due to faithful (non-mutational) replication, and the second describes changes due to mutation. For this reason, equation \eqref{nD_replicator_mutator} is called the \emph{replicator-mutator equation} in the evolutionary game theory literature, where the individual `types' are interpreted to be pure strategies. If in addition, each $w_i(\mathbf{x})$ is linear in $\mathbf{x}$, meaning we can write $w_i(\mathbf{x}) = \sum_{j}a_{ij}x_j$ for some set of constants $a_{ij}$, then we get the replicator-mutator equation for matrix games, and the constants $a_{ij}$ form the `payoff matrix'. As is well-known, the replicator equation (without mutation) for matrix games with $m$ pure strategies is equivalent to the generalized Lotka-Volterra equations for a community with $m-1$ species\citep{hofbauer_evolutionary_1998}, providing the connection to community ecology.  Equation \eqref{nD_replicator_mutator} is also equivalent to Eigen's \emph{quasispecies equation} from molecular evolution if each `type' is interpreted as a genetic sequence and each $w_i(\mathbf{x})$ is a constant function\footnote{Mutational effects are often additionally assumed to act through direct `transmission probabilities' of mutating from one type to another. This means that we can write $Q_i(\mathbf{p}) = \sum\limits_{j\neq i}Q_{ij}p_j$, where $Q_{ij} \geq0$ is a constant describing the probability of a $j \to i$ mutation (conditioned on the occurrence of a mutation). Substituting this into \eqref{nD_replicator_mutator} yields an equation in terms of `$Q$-matrices' or `mutation matrices' that may be more familiar to some biologists.}. We can now calculate how the mean of any `trait level' quantity $f$, defined as $f_i$ for the $i$\textsuperscript{th} trait, changes in the population (For example, setting $f_i = x_i$ gives us the mean trait value in the population). Multiplying both sides of equation \eqref{nD_replicator_mutator} by $f_i$ and summing over all $i$, we obtain

@@

Following ideas similar in spirit to \citep{mcleod_social_2019}, we can also use \eqref{nD_Ito_SDE} to write down stochastic equations for $p_i(t)$, the frequency of type $i$ individuals in the population to contrast with the replicator-mutator equation \eqref{nD_replicator_mutator} derived above. In appendix \ref{App_density_to_freq}, we show that we can use It\^{o}'s lemma to write  down a general stochastic equation for the frequencies of each type in the population.  Unlike \cite{mcleod_social_2019}, we make no assumptions about separations of time scales or strength of selection and are able to present an entirely general calculation. Letting $\tau_i(\mathbf{x}) = b^{(\textrm{int})}_{i}(\mathbf{x}) + d^{(\textrm{int})}_i(\mathbf{x})$ be the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type and denoting $\overline{\tau} = \sum_i p_i\tau_i$, we show in appendix \ref{App_density_to_freq} that when the birth and death rates are defined by \eqref{nD_functional_forms_for_replicator}, then the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{X}_t$, changes according to the equation:

@@

&- \frac{1}{K}\underbrace{\frac{1}{N(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt}_{\substack{\text{Directional finite population effects}\\\text{due to differential turnover rates}}}\\

@@

where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes. The first term in this expression is the same as equation \eqref{nD_replicator_mutator} and describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation. We saw in the previous section that the replicator-mutator equation is generically satisfied in the infinite population limit for all birth-death processes that are in the (very general) functional forms \eqref{nD_functional_forms_for_replicator}, a result that is made explicit in \eqref{nD_eqn_for_frequencies} (just take $K \to \infty$ to convince yourself). The second term is an additional directional force on population composition that acts exactly analogously to the replicator equation but in the opposite direction (A higher relative $\tau_i$ leads to a lower frequency), and furthermore, is particular to finite populations. This term describes a biasing effect due to differential turnover rates and can intuitively be understood in the following sense: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus greater demographic stochasticity. Types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period (notice the minus sign before the second term in \eqref{nD_eqn_for_frequencies}). This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, and this is reflected in the $1/N$ factor in this term. This stochastic analog of selection for reduced turnover rates is the force responsible for the `reversal of the direction of deterministic selection' observed in previous studies \citep{constable_demographic_2016, mcleod_social_2019} and is an effect that is only seen in finite populations. Note that types that tend to increase the \emph{total} population size (such as altruists in evolutionary theory and mutualists in ecological communities) experience a smaller negative effect due to differential turnover rates than types that do not facilitate such an increase (such as cheaters and highly competitive species), which could explain why mutualists and cooperators routinely persist for long periods in finite population models despite extinction being predicted in infinite population models. The fact that the entire term is multiplied by $(KN(t))^{-1}$, however, suggests that the effect of this force is weak for medium to large populations, which explains why the aforementioned persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection + weak mutation limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0 and therefore no longer contributes to the total dynamics, thus uncovering the contributions of the second term.\\
Finally, the last term describes non-directional stochastic effects due to fluctuations and has a `spreading effect'. To the highest order in $\sqrt{K}$, the effect of these fluctuations can be described through the weak noise approximation, as we explain below. However, the $1/K$ factor in the second term of \eqref{nD_eqn_for_frequencies} means that the effect of differential turnover rates disappears in this weak noise regime  (and also in other such approximate treatments, such as the Chemical langevin).

@@

    b_K(x|\psi) &= \mu b_{\textrm{mut}}(x|\psi) + b_{\textrm{int}}(x|\psi)\psi(x,t)\\
    d_K(x|\psi) &= d_{\textrm{int}}(x|\psi)\psi(x,t)

@@

where $\mu \geq 0$ is a mutation rate. Here, $ b_{\textrm{mut}}(x|\psi)$ describes birth due to mutations, $b_{\textrm{int}}(x|\psi)$ describes birth due to ecological interactions with conspecifics, for example, due to mate choice in the sexual case (This function will just take the form of a constant describing intrinsic duplication rate in the asexual case). Similarly, $d_{\textrm{int}}(x|\psi)$ describes death due to interaction with conspecifics, for example due to competition for resources. Substituting equation \eqref{BD_for_kimura} into \eqref{deterministic_traj}, we obtain

>>>>

b^{(K)}_i(\mathbf{x}) &= \mu Q_i(\mathbf{x}) + b^{(\textrm{ind})}_{i}(\mathbf{x})x_i\\
d^{(K)}_i(\mathbf{x}) &= d^{(\textrm{ind})}_i(\mathbf{x})x_i

@@

where $\mu \geq 0$ is a constant describing the mutation rate, $Q_i(\mathbf{x})$ is a real-valued function that describes the birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$, and $b^{(\textrm{ind})}_{i}(\mathbf{x})$ and $d^{(\textrm{ind})}_{i}(\mathbf{x})$ are real-valued functions that respectively describe the per-capita birth and death rate of type $i$ individuals. Our assumptions of the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates. Plugging these definitions into \eqref{nD_det_limit} and writing it down component-wise, we obtain the equation

@@

Given a state $\mathbf{x}(t)$, we can now compute the total (scaled) population size and the frequency of each type in the population as:

@@

We can also calculate the statistical mean value of any type level quantity $f$ in the population as

@@

Where we have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. The first term of \eqref{nD_replicator_mutator} describes changes due to faithful (non-mutational) replication, and the second describes changes due to mutation. For this reason, equation \eqref{nD_replicator_mutator} is called the \emph{replicator-mutator equation} in the evolutionary game theory literature, where the individual `types' are interpreted to be pure strategies. If in addition, each $w_i(\mathbf{x})$ is linear in $\mathbf{x}$, meaning we can write $w_i(\mathbf{x}) = \sum_{j}a_{ij}x_j$ for some set of constants $a_{ij}$, then we get the replicator-mutator equation for matrix games, and the constants $a_{ij}$ form the `payoff matrix'. As is well-known, the replicator equation (without mutation) for matrix games with $m$ pure strategies is equivalent to the generalized Lotka-Volterra equations for a community with $m-1$ species\citep{hofbauer_evolutionary_1998}, providing the connection to community ecology.  Equation \eqref{nD_replicator_mutator} is also equivalent to Eigen's \emph{quasispecies equation} from molecular evolution if each `type' is interpreted as a genetic sequence and each $w_i(\mathbf{x})$ is a constant function\footnote{Mutational effects are often additionally assumed to act through direct `transmission probabilities' of mutating from one type to another. This means that we can write $Q_i(\mathbf{p}) = \sum\limits_{j\neq i}Q_{ij}p_j$, where $Q_{ij} \geq0$ is a constant describing the probability of a $j \to i$ mutation (conditioned on the occurrence of a mutation). Substituting this into \eqref{nD_replicator_mutator} yields an equation in terms of `$Q$-matrices' or `mutation matrices' that may be more familiar to some biologists.}. We can now calculate how the mean of any `type level' quantity $f$, defined as $f_i$ for the $i$\textsuperscript{th} type, changes in the population (For example, if each type is a phenotype for a trait such as height, which can be assigned a numerical value, then setting $f_i = \textit{value of $i\textsuperscript{th}$ phenotype}$ gives us the mean trait value in the population). Multiplying both sides of equation \eqref{nD_replicator_mutator} by $f_i$ and summing over all $i$, we obtain

@@

Following ideas similar in spirit to \citep{mcleod_social_2019}, we can also use \eqref{nD_Ito_SDE} to write down stochastic equations for $p_i(t)$, the frequency of type $i$ individuals in the population to contrast with the replicator-mutator equation \eqref{nD_replicator_mutator} derived above. In appendix \ref{App_density_to_freq}, we show that we can use It\^{o}'s formula to write  down a general stochastic equation for the frequencies of each type in the population. Unlike \citep{mcleod_social_2019}, we make no assumptions about the separation of ecological and evolutionary time scales or the strength of selection and are able to present an entirely general calculation. Letting $\tau_i(\mathbf{x}) = b^{(\textrm{int})}_{i}(\mathbf{x}) + d^{(\textrm{int})}_i(\mathbf{x})$ be the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type and denoting $\overline{\tau} = \sum_i p_i\tau_i$, we show in appendix \ref{App_density_to_freq} that when the birth and death rates are defined by \eqref{nD_functional_forms_for_replicator}, then the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{X}_t$, changes according to the equation:

@@

&- \frac{1}{K}\underbrace{\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt}_{\substack{\text{Directional finite population effects}\\\text{due to differential turnover rates}}}\\

@@

where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes. The first term in this expression is the same as equation \eqref{nD_replicator_mutator} and describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation. We saw in the previous section that the replicator-mutator equation is generically satisfied in the infinite population limit for all birth-death processes that are in the (very general) functional forms \eqref{nD_functional_forms_for_replicator}, a result that is made explicit in \eqref{nD_eqn_for_frequencies} (just take $K \to \infty$ to convince yourself). The second term is an additional directional force on population composition that is only seen in finite populations. Mathematically, this term acts exactly analogously to the replicator equation but in the opposite direction (A higher relative $\tau_i$ leads to a decrease in frequency - notice the minus sign before the second term in \eqref{nD_eqn_for_frequencies}). Biologically, this term describes a biasing effect due to differential turnover rates and can intuitively be understood through the following reasoning: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus experience more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N$ factor in this term. This stochastic analog of selection for reduced turnover rates is the force responsible for the `reversal of the direction of deterministic selection' induced by demographic noise in previous studies \citep{constable_demographic_2016, mcleod_social_2019}. Note that types that tend to increase the \emph{total} population size $N(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase (such as cheaters and highly competitive species), which could explain why the former routinely persist for long periods in finite population models despite extinction being predicted in infinite population models. The fact that the entire term is multiplied by $(KN(t))^{-1}$ suggests that the effect of this force is weak for medium to large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection + weak mutation limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0. It therefore no longer contributes to the total dynamics and allows us to see the contributions of the second term.\\
Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect'. To the highest order in $\sqrt{K}$, the effect of these fluctuations can be described through the weak noise approximation, as we explain below. However, the $1/K$ factor in the second term of \eqref{nD_eqn_for_frequencies} means that the effect of differential turnover rates disappears in this weak noise regime  (and also in other such approximate treatments, such as the Chemical Langevin equation).

@@

b_K(x|\psi) &= \mu b_{\textrm{mut}}(x|\psi) + b_{\textrm{ind}}(x|\psi)\psi(x,t)\\
d_K(x|\psi) &= d_{\textrm{ind}}(x|\psi)\psi(x,t)

@@

Here, $ b_{\textrm{mut}}(x|\psi)$ describes birth due to mutations, and $\mu \geq 0$ is a mutation rate. The functions $b_{\textrm{ind}}(x|\psi)$ and $d_{\textrm{ind}}(x|\psi)$ describe the per-capita birth rate and death rate of type $x$ in a population $\psi$. For example,  $b_{\textrm{ind}}(x|\psi)$ may incorporate the effects of mate choice in the sexual case or intrinsic duplication rate in the asexual case, and $d_{\textrm{int}}(x|\psi)$ may model death due to intraspecific competition for resources. Substituting equation \eqref{BD_for_kimura} into \eqref{deterministic_traj}, we obtain

### 2023-02-04 ind_files/99_Appendix_B.tex
Aesthetic changes to .cls file - make title bold in certificate

We first recall the version of the multi-dimensional It\^{o}'s lemma that will be relevant to us. Consider an $m$-dimensional real It\^{o} process $\mathbf{X}_t$ given by the solution to

@@

where $\boldsymbol{\mu}: \mathbb{R}^m \to \mathbb{R}^m$ is the `drift vector' and $\boldsymbol{\sigma}: \mathbb{R}^{m} \to \mathbb{R}^{m \times m}$ is the `diffusion matrix'. Let $f: \mathbb{R}^m \to \mathbb{R}$ be an arbitrary $C^2(\mathbb{R}^m)$ function. Then, It\^{o}'s lemma (\cite{oksendal_stochastic_1998}, Section 4.2) states that  the stochastic process $f(\mathbf{X}_t)$ must satisfy:

@@

\label{nD_Ito_lemma}

@@

Thus, $f_i(\mathbf{X}_t)$ gives us the frequency of type $i$ individuals when the population is described by the vector $\mathbf{X}_t$. This function is obviously $C^2(\mathbb{R}^m)$, and we can thus use It\^{o}'s lemma \eqref{nD_Ito_lemma} to describe how it changes over time. The $j\textsuperscript{th}$ element of the gradient of $f_i$ is given by:

@@

where we have defined the total population size $N = \sum_i x_i$ and the frequency of the $i\textsuperscript{th}$ type $p_i = f_i(x)$ to reduce clutter, and used the fact that $\frac{\partial x_j}{\partial x_k} = \delta_{jk}$. The $jk\textsuperscript{th}$ element of the Hessian is given by:

@@

Thus, for the first term of \eqref{nD_Ito_lemma}, we have:

@@

If the birth and death rates take the form of \eqref{nD_functional_forms_for_replicator}, then it is easy to see using \eqref{nD_det_limit_fitess_defn} that equation \eqref{nD_for_Ito_first_term} is exactly the RHS of \eqref{nD_replicator_intermediate_1}. Thus, when the birth and death rates take the form of \eqref{nD_functional_forms_for_replicator}, then \eqref{nD_for_Ito_first_term} describes the deterministic component of the dynamics as described by the replicator-mutator equation, Price equation, etc. in the infinite population limit. These are the effects of selection and mutation at the infinite population limit. However, the finiteness of the population adds a second term to these dynamics, described by the second term that multiplies $dt$ in \eqref{nD_Ito_lemma}. To calculate it, we first calculate:

@@

and thus, the second term multiplying $dt$ in \eqref{nD_Ito_lemma} is given by:

@@

Thus, using \eqref{nD_jacobian_for_ito}, we see that the last term on the RHS of \eqref{nD_Ito_lemma} is given by

@@

Putting equations \eqref{nD_for_Ito_first_term}, \eqref{nD_for_Ito_second_term} and \eqref{nD_for_Ito_third_term} into \eqref{nD_Ito_lemma}, we see that $p_i = f_i(\mathbf{X})_t$, the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{X}_t$, changes according to the equation:

>>>>

We first recall the version of the multi-dimensional It\^{o}'s formula that will be relevant to us. Consider an $m$-dimensional real It\^{o} process $\mathbf{X}_t$ given by the solution to

@@

where $\boldsymbol{\mu}: \mathbb{R}^m \to \mathbb{R}^m$ is the `drift vector' and $\boldsymbol{\sigma}: \mathbb{R}^{m} \to \mathbb{R}^{m \times m}$ is the `diffusion matrix'. Let $f: \mathbb{R}^m \to \mathbb{R}$ be an arbitrary $C^2(\mathbb{R}^m)$ function. Then, It\^{o}'s formula (\cite{oksendal_stochastic_1998}, Section 4.2) states that  the stochastic process $f(\mathbf{X}_t)$ must satisfy:

@@

\label{nD_Ito_formula}

@@

Thus, $f_i(\mathbf{X}_t)$ gives us the frequency of type $i$ individuals when the population is described by the vector $\mathbf{X}_t$. This function is obviously $C^2(\mathbb{R}^m)$, and we can thus use It\^{o}'s formula \eqref{nD_Ito_formula} to describe how it changes over time. The $j\textsuperscript{th}$ element of the gradient of $f_i$ is given by:

@@

where we have defined the total (scaled) population size\footnote{This is $N_{K}(t)$ in the main text, but we omit the subscript here to reduce notational clutter}  $N = \sum_i x_i$ and the frequency of the $i\textsuperscript{th}$ type $p_i = f_i(x)$ and used the fact that $\frac{\partial x_j}{\partial x_k} = \delta_{jk}$. The $jk\textsuperscript{th}$ element of the Hessian is given by:

@@

Thus, for the first term of \eqref{nD_Ito_formula}, we have:

@@

If the birth and death rates take the form of \eqref{nD_functional_forms_for_replicator}, then it is easy to see using \eqref{nD_det_limit_fitess_defn} that equation \eqref{nD_for_Ito_first_term} is exactly the RHS of \eqref{nD_replicator_intermediate_1}. Thus, when the birth and death rates take the form of \eqref{nD_functional_forms_for_replicator}, then \eqref{nD_for_Ito_first_term} describes the deterministic component of the dynamics as described by the replicator-mutator equation, Price equation, etc. in the infinite population limit. These are the effects of selection and mutation at the infinite population limit. However, the finiteness of the population adds a second term to these dynamics, described by the second term that multiplies $dt$ in \eqref{nD_Ito_formula}. To calculate it, we first calculate:

@@

and thus, the second term multiplying $dt$ in \eqref{nD_Ito_formula} is given by:

@@

Thus, using \eqref{nD_jacobian_for_ito}, we see that the last term on the RHS of \eqref{nD_Ito_formula} is given by

@@

Putting equations \eqref{nD_for_Ito_first_term}, \eqref{nD_for_Ito_second_term} and \eqref{nD_for_Ito_third_term} into \eqref{nD_Ito_formula}, we see that $p_i = f_i(\mathbf{X})_t$, the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{X}_t$, changes according to the equation:

### 2023-02-04 main.tex
Add section on stochastic trait freq dynamics
Aesthetic changes to .cls file - make title bold in certificate

\usepackage[toc]{appendix}
\setcounter{tocdepth}{3}

@@

\title{Eco-evolutionary dynamics of finite populations from demographic first principles}

@@

\thesisabstract{
Some central questions of ecology and evolution ask how many distinct variants of an entity can emerge and coexist in sympatry. Examples include the study of diversity in standing genetic variation (alleles), polymorphisms (genotypes/phenotypes), and sympatric speciation (species). Historically, these questions have often been studied through phenomenological models formulated for infinitely large populations. Stochastic individual-based models (IbMs), where ecological rules are specified for each individual, are more realistic but are rarely studied analytically.
In this thesis, I present a review and synthesis of how ideas from statistical physics can be used to analytically formulate and study IbMs from biological first principles. Starting from a stochastic ‘birth-death process’, I show how density dependence of the population dynamics can be used to obtain a continuous approximation of the system. I then show how well-known equations such as the replicator-mutator and Price equations are recovered in the infinite population limit. The dynamics of large, finite populations can be expressed as stochastic deviations from the deterministic result predicted by these well-known models using a ‘weak noise approximation’. As an application, I show how this formalism can be used to describe a version of the logistic equation for finite populations. Finally, I extend this theory to study quantitative traits through a stochastic field equation. In the infinite population limit, I recover well-known models such as Kimura's continuum-of-alleles and the canonical equation of adaptive dynamics. As an application, I show how to study finite-population effects on evolutionary branching, a process considered important for sympatric speciation.
}

@@

\begin{document}
\thesisfront
\listoftables
\listoffigures	
\chapter{Introduction}
\input{ind_files/01_Introduction}
\chapter{Adaptive diversification: a crash course}\label{chap_AD}
\input{ind_files/02_Foundations}
\chapter{Population dynamics from stochastic first principles}\label{chap_BD}
\input{ind_files/03_BD_processes.tex}
\chapter{Examples and Applications}\label{chap_examples}
\input{ind_files/04_polymorphisms}
% \chapter{}
% \input{ind_files/chapter_03}
% \chapter{} <-- uncomment and add as required
% \input{ind_files/chapter_04}

\begin{appendices}
\chapter{Relation between Ito SDEs and Fokker-Planck equations}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\end{appendices}
\printbibliography[title=References]
\end{document}

>>>>

\setcounter{tocdepth}{3}

@@

\title{Eco-evolutionary dynamics of finite populations from first principles}

@@

\thesisabstract{
Some central questions of ecology and evolution ask how many distinct variants of an entity can emerge and coexist in sympatry. Examples include the study of standing genetic variation (alleles), polymorphisms (genotypes/phenotypes), and sympatric speciation (species). Historically, these questions have often been studied through phenomenological models formulated for infinitely large populations. Stochastic individual-based models (IbMs), where mechanistic rules are specified for each individual, are more realistic but are rarely studied analytically.\\
\\
In this thesis, I show how ideas from statistical physics can be used to analytically formulate and study IbMs from biological first principles. Starting from a stochastic ‘birth-death process’, I show how density dependence of the population dynamics can be used to obtain a continuous approximation of the system as a stochastic differential equation (SDE). I show that well-known ‘fundamental theorems’ of evolution, such as the Price equation, are recovered in the infinite population limit. My stochastic formulation reveals a new directional evolutionary force in finite populations that lends itself to a simple biological interpretation and has no analogue in infinitely large populations. As an application, I show how this formalism can be used to study cooperation in finite populations. Finally, I extend this theory to study quantitative traits through a stochastic field equation, which can be thought of as an infinite-dimensional SDE. In the infinite population limit, I recover well-known models such as Kimura's continuum-of-alleles and adaptive dynamics. My work thus highlights the connections between various equations in ecology and evolution by revealing their common stochastic underpinnings.
}

@@

\begin{document}
\thesisfront
\listoftables
\listoffigures	
\chapter{Introduction}
\input{ind_files/01_Introduction}
\chapter{Adaptive diversification: a crash course}\label{chap_AD}
\input{ind_files/02_Foundations}
\chapter{Population dynamics from stochastic first principles}\label{chap_BD}
\input{ind_files/03_BD_processes.tex}
\chapter{Examples and Applications}\label{chap_examples}
\input{ind_files/04_polymorphisms}
% \chapter{}
% \input{ind_files/chapter_05}
% \chapter{} <-- uncomment and add as required
% \input{ind_files/chapter_06}

\begin{appendices}
\chapter{Deriving the Fokker-Planck equations for It\^{o} SDEs}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\end{appendices}

\printbibliography[title=References]
\end{document}
